### 2016-05-27 Paper_v2.tex
Materials and methods done

Rate processes were written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$) in the complement model.
The kinetic term for the formation of C4a, C4b, C2a and C2b, lectin pathway activation, and C3 and C5 convertase activity was given by:
\begin{equation}\label{eqn:rate-saturation}
	\bar{r}_{j} = k_{j}^{max}\epsilon_{i}\left(\frac{x_{s}^{\eta}}{K_{js}^{\eta} + x_{s}^{\eta}}\right)
\end{equation}
where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the abundance of the enzyme catalyzing reaction $j$,
$\eta$ denotes a cooperativity parameter, and $K_{js}$ denotes the saturation constant for species $s$ in reaction $j$.
We used mass action kinetics to model protein-protein binding interactions within the network:
\begin{equation}\label{eqn:rate-action}
	\bar{r}_{j} = k_{j}^{max}\prod_{s\in{m_{j}^{-}}}x_{s}^{-\sigma_{sj}}
\end{equation}
where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\sigma_{sj}$ denotes the stoichiometric coefficient for species $s$ in reaction $j$,
and $s\in{m_{j}}$ denotes the set of \textit{reactants} for reaction $j$.
The control terms $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
For each rate, we used a rule-based approach to select from competing control factors.
If rate j was influenced by $1,\dots,m$ factors, we modeled this relationship as
$v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
variable. Each regulatory transfer function took the form:
\begin{equation}\label{eqn:control-factor}
	f_{ij}\left(\mathcal{Z}_{i},k_{ij},\eta_{ij}\right)=k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}/\left({1 + k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}}\right)
\end{equation}where $\mathcal{Z}_{i}$ denotes the abundance of factor $i$, $k_{ij}$ denotes a gain parameter, and $\eta_{ij}$ denotes a cooperativity parameter.
In this study, we used $\mathcal{I}_{j}\in\left\{mean\right\}$ \cite{pr3010178}. If a process has no modifying factors, $v_{j}=1$.
The model equations implemented in MATLAB and solved using the \texttt{ODE15s} routine (The Mathworks, Natick MA).
The complement model code and parameter ensemble can be downloaded from http://www.varnerlab.org.

\subsection*{Estimation of an ensemble of model parameters.}
Model parameters were estimated by minimizing the difference between simulations and experimental C3a and C5a measurements (squared residual):
\begin{equation}\label{eqn:objective-function}
	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated
value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is with respect to
time, while the inner summation is with respect to state.

We minimized the model residual using Dynamic Optimization with Particle Swarms (DOPS). DOPS is a novel metaheuristic that combines multi swarm particle swarm optimization (PSO) with a greedy global optimization algorithm called dynamically dimensioned search (DDS). DOPS is much faster than conventional global optimizers and has the ability to find near optimal solutions for high dimensional systems within a relatively few function evaluations. It uses an adaptive switching strategy based on error convergence rates to switch from swarms search to DDS search. This enables it to find quickly, globally optimal or close to globally optimal solutions even in the presence of many local minima. In the swarm search, for each iteration the particles compute error within each sub-swarm by evaluating the model equations using their specific parameter vector realization. From each of these points within a sub-swarm a local best is identified. This along with the particle best within the sub-swarm $\mathcal{S}_{k}$ is used to update the parameter estimate for each particle using the following rules:

\begin{equation}\label{eqn:update-rule}
	%\mathbf{z}_{i,j} = \theta_{1,j-1}\mathbf{z}_{i,j-1} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{z}_{i,j-1}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G}_{k} - \mathbf{z}_{i,j-1}\right)
	{z}_{i,j} = \theta_{1,j-1}{z}_{i,j-1} + \theta_{2}{r}_{1}\left(\mathcal{L}_{i} - {z}_{i,j-1}\right) + \theta_{3}{r}_{2}\left(\mathcal{G}_{k} - {z}_{i,j-1}\right)
\end{equation}
where ${z}_{i,j}$ is the parameter vector, $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ were adjustable parameters, $\mathcal{L}_{i}$ denotes the best solution found by particle $i$ within sub-swarm
$\mathcal{S}_{k}$ for function evaluations $1\rightarrow j-1$, and
$\mathcal{G}_{k}$ denotes the best solution found over all particles within sub-swarm $\mathcal{S}_{k}$.
The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model parameters ($\mathcal{K}\times{1}$).
At the conclusion of the swarm phase, the overall best particle, $\mathcal{G}_{k}$, over the k sub-swarms was used to initialize the DDS phase.
For the DDS phase, the best parameter estimate was updated using the rule:
\begin{equation}
  \mathcal{G}_{new}({J})=\begin{cases}
    \mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J}), & \text{if $\mathcal{G}_{new}(\mathbf{J})<\mathcal{G}(\mathbf{J})$}.\\
    \mathcal{G}(\mathbf{J}), & \text{otherwise}.
  \end{cases}
\end{equation}
where $\mathbf{J}$ is a vector representing the subset of dimensions that are being perturbed, $\mathbf{r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$,
and $\sigma$ denotes the perturbation amplitude:
\begin{equation}
	\sigma = {R}(\mathbf{p}^U - \mathbf{p}^L)
\end{equation}
where ${R}$ is the scalar perturbation size parameter, $\mathbf{p}^U$ and $\mathbf{p}^L$ are ($\mathcal{K}\times{1}$) vectors that represent the maximum and minimum bounds on each dimension. The set $\mathbf{J}$ was constructed using a monotonically decreasing probability function $\mathcal{P}_{i}$ that represents a threshold for determining whether a specific dimension $j$ was perturbed or not. DDS updates are greedy; $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than $\mathcal{G}$.
At the end of DDS phase we obtain the optimal vector  $\mathcal{G}$ for our model which we use for plotting best fits against the experimental data. We perturb this parameter vector to generate an ensemble of parameter vectors. The quality of parameter estimates was measured using goodness of fit (model residual).  The DOPS routine was implemented in the MATLAB programming language.

@@

\subsection*{Global sensitivity analysis of model performance}
We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which parameters controlled the performance of the reduced order model \cite{SOBOL_METHOD}. We computed the total sensitivity index of each parameter relative to four performance objectives, each objective was based on the sum of squared errors between model and experimental data for C3a alternate, C5a alternate, C3a lectin, and C5a lectin simulations. We established the sampling bounds for each parameter from the minimum and maximum value of that parameter in the parameter set ensemble.
We used the sampling method of Saltelli \textit{et al.} \cite{saltelli2010variance} to compute a family of $N\left(2d+2\right)$ parameter sets which obeyed our parameter ranges,
where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, $N$ = 200 and $d$ = 28, so the total sensitivity indices were computed from
11,600 model evaluations. The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \cite{SALIB}.

>>>>

Rate processes were written as the product of a kinetic term ($\bar{r}_{j}$) and a control term ($v_{j}$) in the complement model.
The kinetic term for the formation of C4a, C4b, C2a and C2b, lectin pathway activation, and C3 and C5 convertase activity was given by:
\begin{equation}\label{eqn:rate-saturation}
	\bar{r}_{j} = k_{j}^{max}\epsilon_{i}\left(\frac{x_{s}^{\eta}}{K_{js}^{\eta} + x_{s}^{\eta}}\right)
\end{equation}
where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the abundance of the enzyme catalyzing reaction $j$,
$\eta$ denotes a cooperativity parameter, and $K_{js}$ denotes the saturation constant for species $s$ in reaction $j$.
We used mass action kinetics to model protein-protein binding interactions within the network:
\begin{equation}\label{eqn:rate-action}
	\bar{r}_{j} = k_{j}^{max}\prod_{s\in{m_{j}^{-}}}x_{s}^{-\sigma_{sj}}
\end{equation}
where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\sigma_{sj}$ denotes the stoichiometric coefficient for species $s$ in reaction $j$,
and $s\in{m_{j}}$ denotes the set of \textit{reactants} for reaction $j$.
The control terms $0\leq v_{j}\leq 1$ depended upon the combination of factors which influenced rate process $j$.
For each rate, we used a rule-based approach to select from competing control factors.
If rate j was influenced by $1,\dots,m$ factors, we modeled this relationship as
$v_{j}=\mathcal{I}_{j}\left(f_{1j}\left(\cdot\right),\hdots,f_{mj}\left(\cdot\right)\right)$
where $0\leq f_{ij}\left(\cdot\right)\leq 1$ denotes a regulatory transfer function quantifying the influence of factor $i$ on rate $j$.
The function $\mathcal{I}_{j}\left(\cdot\right)$ is an integration rule which maps the output of regulatory transfer functions into a control
variable. Each regulatory transfer function took the form:
\begin{equation}\label{eqn:control-factor}
	f_{ij}\left(\mathcal{Z}_{i},k_{ij},\eta_{ij}\right)=k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}/\left({1 + k_{ij}^{\eta_{ij}}\mathcal{Z}_{i}^{\eta_{ij}}}\right)
\end{equation}where $\mathcal{Z}_{i}$ denotes the abundance of factor $i$, $k_{ij}$ denotes a gain parameter, and $\eta_{ij}$ denotes a cooperativity parameter.
In this study, we used $\mathcal{I}_{j}\in\left\{min,max\right\}$ \cite{pr3010178}. If a process has no modifying factors, $v_{j}=1$.
The model equations were implemented in MATLAB and solved using the \texttt{ODE15s} routine (The Mathworks, Natick MA).
The complement model code and parameter ensemble can be downloaded from http://www.varnerlab.org.

\subsection*{Estimation of an ensemble of model parameters.}
We minimized the residual between simulations and experimental C3a and C5a measurements using Dynamic Optimization with Particle Swarms (DOPS). DOPS minimized the objective:
\begin{equation}\label{eqn:objective-function}
	\min_{\mathbf{k}} \sum_{\tau=1}^{\mathcal{T}}\sum_{j=1}^{\mathcal{S}}\left(\frac{\hat{x}_{j}\left(\tau\right) - x_{j}\left(\tau,\mathbf{k}\right)}{\omega_{j}\left(\tau\right)}\right)^{2}
\end{equation}where $\hat{x}_{j}\left(\tau\right)$ denotes the measured value of species $j$ at time $\tau$, $x_{j}\left(\tau,\mathbf{k}\right)$ denotes the simulated
value for species $j$ at time $\tau$, and $\omega_{j}\left(\tau\right)$ denotes the experimental measurement variance for species $j$ at time $\tau$. The outer summation is with respect to
time, while the inner summation is with respect to state. DOPS is a novel metaheuristic that combines multi swarm particle swarm optimization (PSO) with a greedy global optimization algorithm called dynamically dimensioned search (DDS).
DOPS is faster than conventional global optimizers and has the ability to find near optimal solutions for high dimensional systems within a relatively few function evaluations.
It uses an adaptive switching strategy based on error convergence rates to switch from the particle swarm to DDS search phases.
This enables DOPS to quickly estimate globally optimal or near optimal solutions even in the presence of many local minima.
In the swarm search, for each iteration the particles compute error within each sub-swarm by evaluating the model equations using their specific parameter vector realization. From each of these points within a sub-swarm a local best is identified. This along with the particle best within the sub-swarm $\mathcal{S}_{k}$ is used to update the parameter estimate for each particle using the following rules:

\begin{equation}\label{eqn:update-rule}
	%\mathbf{z}_{i,j} = \theta_{1,j-1}\mathbf{z}_{i,j-1} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{z}_{i,j-1}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G}_{k} - \mathbf{z}_{i,j-1}\right)
	{z}_{i,j} = \theta_{1,j-1}{z}_{i,j-1} + \theta_{2}{r}_{1}\left(\mathcal{L}_{i} - {z}_{i,j-1}\right) + \theta_{3}{r}_{2}\left(\mathcal{G}_{k} - {z}_{i,j-1}\right)
\end{equation}
where ${z}_{i,j}$ is the parameter vector, $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ were adjustable parameters, $\mathcal{L}_{i}$ denotes the best solution found by particle $i$ within sub-swarm
$\mathcal{S}_{k}$ for function evaluations $1\rightarrow j-1$, and
$\mathcal{G}_{k}$ denotes the best solution found over all particles within sub-swarm $\mathcal{S}_{k}$.
The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model parameters ($\mathcal{K}\times{1}$).
At the conclusion of the swarm phase, the overall best particle, $\mathcal{G}_{k}$, over the k sub-swarms was used to initialize the DDS phase.
For the DDS phase, the best parameter estimate was updated using the rule:
\begin{equation}
  \mathcal{G}_{new}({J})=\begin{cases}
    \mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J}), & \text{if $\mathcal{G}_{new}(\mathbf{J})<\mathcal{G}(\mathbf{J})$}.\\
    \mathcal{G}(\mathbf{J}), & \text{otherwise}.
  \end{cases}
\end{equation}
where $\mathbf{J}$ is a vector representing the subset of dimensions that are being perturbed, $\mathbf{r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$,
and $\sigma$ denotes the perturbation amplitude:
\begin{equation}
	\sigma = {R}(\mathbf{p}^U - \mathbf{p}^L)
\end{equation}
where ${R}$ is the scalar perturbation size parameter, $\mathbf{p}^U$ and $\mathbf{p}^L$ are ($\mathcal{K}\times{1}$) vectors that represent the maximum and minimum bounds on each dimension. The set $\mathbf{J}$ was constructed using a monotonically decreasing probability function $\mathcal{P}_{i}$ that represents a threshold for determining whether a specific dimension $j$ was perturbed or not. DDS updates are greedy; $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than $\mathcal{G}$.
At the end of DDS phase we obtain the optimal vector  $\mathcal{G}$ for our model which we use for plotting best fits against the experimental data. We perturb this parameter vector to generate an ensemble of parameter vectors. The quality of parameter estimates was measured using goodness of fit (model residual).
The DOPS routine was implemented in MATLAB (The Mathworks, Natick MA) and can be downloaded from http://www.varnerlab.org.

@@

\subsection*{Global sensitivity analysis of model performance.}
We conducted a global sensitivity analysis to estimate which parameters controlled the performance of the reduced order model using the Sobol method \cite{SOBOL_METHOD}.
We computed the total sensitivity index of each parameter relative to the training residual for the C3a alternate, C5a alternate, C3a lectin, and C5a lectin cases.
We established the sampling bounds for each parameter from the minimum and maximum value for that parameter in the parameter ensemble.
We used the sampling method of Saltelli \textit{et al.} to compute a family of $N\left(2d+2\right)$ parameter sets which obeyed our parameter ranges,
where $N$ was the number of trials, and $d$ was the number of parameters in the model \cite{saltelli2010variance}.
In our case, $N$ = 200 and $d$ = 28, so the total sensitivity indices were computed from 11,600 model evaluations.
The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \cite{SALIB}.
